---
title: 'HultPrize2018: Electric Power Loss & Covariates'
author: "C. McClintock"
date: "Spring 2018"
output: 
  html_document: 
    toc: TRUE
    toc_float: TRUE
    depth: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE, echo = FALSE)
setwd("/Users/charmed33/R/HultPrize2018")
library(tidyverse)
library(knitr)
library(dplyr)
```

```{r set, echo = FALSE}
# read in the data, all from the world bank
powerloss <- read_csv("electric_loss.csv", skip = 4) 
# Electric power transmission and distribution losses (% of output)
access <- read_csv("access.csv", skip = 4) 
# Access to electricity (% of population)
population <- read_csv("population.csv", skip = 4)
# Population, total
kwh <- read_csv("kwhpercap.csv", skip = 4)
# Electric power consumption (kWh per capita)
gdp <- read_csv("gdp.csv", skip = 4)
# GDP (current US$)
countries <- read_csv("wb_countrydata.csv")
# Children out of school, female (% of female primary school age)
girlsed <- read_csv("girlsed.csv", skip = 4)

# change some names, use data from 2014, last available
colnames(powerloss)[colnames(powerloss)=="Country Name"] <- "country"
colnames(powerloss)[colnames(powerloss)=="Country Code"] <- "country_code"
colnames(powerloss)[colnames(powerloss)=="2014"] <- "percentloss"

colnames(population)[colnames(population)=="Country Name"] <- "country"
colnames(population)[colnames(population)=="Country Code"] <- "country_code"
colnames(population)[colnames(population)=="2014"] <- "population"

colnames(gdp)[colnames(gdp)=="Country Name"] <- "country"
colnames(gdp)[colnames(gdp)=="Country Code"] <- "country_code"
colnames(gdp)[colnames(gdp)=="2014"] <- "gdp"

colnames(kwh)[colnames(kwh)=="Country Name"] <- "country"
colnames(kwh)[colnames(kwh)=="Country Code"] <- "country_code"
colnames(kwh)[colnames(kwh)=="2014"] <- "kwh"

colnames(access)[colnames(access)=="Country Name"] <- "country"
colnames(access)[colnames(access)=="Country Code"] <- "country_code"
colnames(access)[colnames(access)=="2014"] <- "access"

colnames(girlsed)[colnames(girlsed)=="Country Name"] <- "country"
colnames(girlsed)[colnames(girlsed)=="Country Code"] <- "country_code"
colnames(girlsed)[colnames(girlsed)=="2014"] <- "girlsed"

names(countries) <- c("country_code","region","income_group","country")


# select the 2014 data
powerloss <- select(powerloss, country, country_code, percentloss)
population <- select(population, country, country_code, population)
gdp <- select(gdp, country, country_code, gdp)
kwh <- select(kwh, country, country_code, kwh)
access <- select(access, country, country_code, access)
girlsed <- select(girlsed, country, country_code, girlsed)

# merge it all together
power <- left_join(powerloss, countries, by = 
           c("country_code" = "country_code","country" = "country"))
power <- left_join(power, population, by = 
           c("country_code" = "country_code", "country" = "country"))
power <- left_join(power, gdp, by = 
           c("country_code" = "country_code", "country" = "country"))
power <- left_join(power, kwh, by = 
           c("country_code" = "country_code", "country" = "country"))
power <- left_join(power, access, by = 
           c("country_code" = "country_code", "country" = "country"))
power <- left_join(power, girlsed, by = 
           c("country_code" = "country_code", "country" = "country"))

# create gdp per capita by dividing gdp by population
power <- mutate(power, gdp_percap = gdp / population)
power <- select(power, country, country_code, region, 
                income_group, gdp, gdp_percap, population, everything())
power <- power[complete.cases(power[ , 3]),]
power$income_group <- as.factor(power$income_group)
power <- mutate(power,
                income_group=fct_relevel(income_group,
                  "Low income",
                  "Lower middle income",
                  "Upper middle income",
                  "High income"))
```


## Objective
> Explore trends related to global electricity theft to better understand energy consumption and correlated variables to pitch a scalable, sustainable social enterprise to bring energy to 10 million people currently living in energy poverty. 

## Background
Electricity theft, also called non-technical line loss, costs the energy utility industry $96 billion per year, most of that concentrated in emerging markets. [1](https://www.prnewswire.com/news-releases/96-billion-is-lost-every-year-to-electricity-theft-300453411.html) Of the total theft detected, 80% comes from residential buildings and 20% from commercial and industrial premises. [2](http://www.ajer.org/papers/v4(01)/I0401059069.pdf) Hook lines are the most common used method of electricity theft, with 80% of global power theft is by direct tapping from the main power line. [3](http://file.scirp.org/Html/1-1770179_70365.htm) Theft of electricity causes significant harm to the economy, including include loss of revenue and profit of electricity providers and decrease of the stability and reliability of electrical power grids, as in some countries theft may range up to 40% of the total electricity distributed. [4](https://www.atlantis-press.com/journals/ijcis/25872435) India alone loses $16.2 billion  per year, and plans to invest $21.6 billion  between 2015 and 2025 to develop a smart grid to prevent fraud and theft. [5](http://www.tdworld.com/smart-grid/india-spend-216-billion-smart-grid-infrastructure-2025)

## Data Origin

Data accessed through the [World Bank Open Data Portal](https://data.worldbank.org).

* Data Sets:
    + [Population, total](https://data.worldbank.org/indicator/SP.POP.TOTL)
     + [GDP (current US$)](https://data.worldbank.org/indicator/NY.GDP.MKTP.CD)
      + [Access to electricity (% of population)](https://data.worldbank.org/indicator/EG.ELC.ACCS.ZS)
    + [Electric power consumption (kWh per capita)](https://data.worldbank.org/indicator/EG.USE.ELEC.KH.PC)
     + [Electric power transmission and distribution losses (% of output)](https://data.worldbank.org/indicator/EG.ELC.LOSS.ZS)
     + [Children out of school, female (% of female primary school age)](https://data.worldbank.org/indicator/SE.PRM.UNER.FE.ZS?view=chart)
     
## Exploratory Data Analysis

We are interested in what variables affect and are correlated with theft of electricity, studied through electric power and distribution losses. The variables included in our analysis are electric power transmission and distribution losses (by % of output), access to electricity (% of population), total population, GDP (total US$), and development indicators in GDP per capita, and education of girls by % of female primary school age children out of school.

### Highest percent losses

```{r desc}
arb <- select(power, country, percentloss)
kable(head(arrange(arb, -percentloss), 10), col.names = c("Country", 
                     "Electric power transmission and distribution losses (% of output)"))
```

### Lowest kWh per capita 

```{r desc2}
arb2 <- select(power, country, kwh)
kable(head(arrange(arb2, kwh), 10), col.names = c("Country", "Electric power consumption (kWh per capita)"))
```

### Average Power Loss by Income Group

```{r tab}
percentbyincome <- as.data.frame(power %>% 
             group_by(income_group) %>% 
              summarise(mean(percentloss, na.rm = TRUE)))
kable(percentbyincome, col.names = c("Income Group",
                                     "Average electric power transmission and distribution losses (% of output)"))
```

***

### GDP per Capita vs. kWh per Capita

 First, economic development, measured by GDP per capita and energy use, measured by kWh per capita, are highly correlated in a basic linear model. Interestingly, Iceland has the highest kWh per capita, twice as high as the next country, and will be excluded from the models for clarity.
  
```{r plot 1, echo = FALSE}

ggplot(data = power, mapping = aes(gdp_percap, kwh), na.rm = TRUE) + 
  geom_point(aes(colour = factor(income_group)), size = 2) + 
  geom_smooth(method = "lm", se = FALSE) + coord_cartesian(xlim = c(0,125000), 
  ylim=c(0,25000)) + labs(colour = "Income Level", 
       title = "GDP per Capita vs. kWh per Capita", 
       caption = "based on most recent available data from the World Bank (2014)",
       x = "GDP per Capita", 
       y = "kWh per Capita")
```

***

### GDP per Capita vs. Transmission & Distribution Loss (% of output)

Next, looking at electric power distribution and transmission loss as a function of GDP per capita. As expected, the highest output losses occur in the lowest income countries.  

```{r plot 2, echo = FALSE}
# gdp per capita vs. percent loss
ggplot(data = power, mapping = aes(gdp_percap, percentloss, na.rm = TRUE)) + 
  geom_point(aes(colour = factor(income_group)), size = 2) + geom_smooth(se = FALSE) +
  coord_cartesian(xlim = c(0, 125000)) + labs(colour = "Income Level", 
       title = "GDP per Capita vs. Transmission & Distribution Loss (% of output)", 
       caption = "based on most recent available data from the World Bank (2014)",
       x = "GDP per Capita (US$)", 
       y = "Electric power losses (% of output)")
```

***

### kWh per Capita vs. Transmission & Distribution Loss (% of output)

Low electric power consumption (kWh per capita) is associated with high rates of loss (theft), and high electric power consumer means almost no theft. 

```{r plot 3, echo = FALSE}
# kwh per capita vs. percent loss
ggplot(data = power, mapping = aes(kwh, percentloss, na.rm = TRUE)) + 
  geom_point(aes(colour = factor(income_group)), size = 2) + geom_smooth(se = FALSE) +
  coord_cartesian(xlim = c(0, 25000)) + labs(colour = "Income Level", 
       title = "kWh per Capita vs. Transmission & Distribution Loss (% of output)", 
       caption = "based on most recent available data from the World Bank (2014)",
       x = "Electric power consumption (kWh per capita)", 
       y = "Electric power losses (% of output)")
```

***

### Percent Access to Electricity vs. kWh per Capita

This is the most interesting trend discovered. Power consumption in kWh per capita does not rise substantially until a country has 100% electrification, and upper middle or high income levels. 

```{r graph}
# kwh per capita vs. percent access
ggplot(data = power, mapping = aes(access, kwh), na.rm = TRUE) + 
  geom_point(aes(colour = factor(income_group)), size = 2) + 
  coord_cartesian(ylim = c(0,25000)) + 
  labs(colour = "Income Level", 
       title = "Percent Access to Electricity vs. kWh per Capita", 
       caption = "based on most recent available data from the World Bank (2014)",
       x = "Access to Electricity(%)", 
       y = "Electric power consumption (kWh per capita)")
  
```



## Conclusion

Causation cannot be drawn from correlation, but the results are informative. Total electric power consumption only substantially increases when 100% of the population can be connected to the grid. There is substantial space in the market for investment in theft-resistant technologies.






